---
title: "Mantua Fire Department Payroll"
format:
  html:
    toc: false
    theme: [flatly]
    self-contained: true
    echo: false
    message: false
    warning: false
params:
  start_date: "2026-01-01"
  end_date:   "2026-03-31"
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(DBI)
library(RMariaDB)
library(logger)
library(lubridate)
library(scales)
library(gt)
library(ggplot2)
library(ggrepel)
library(firecore)

CON <- tryCatch({
  DBI::dbConnect(
    drv      = RMariaDB::MariaDB(),
    user     = Sys.getenv("DB_USER"),
    password = Sys.getenv("DB_USER_PASSWORD"),
    host     = Sys.getenv("DB_HOST"),
    port     = 3306,
    dbname   = Sys.getenv("DB_NAME"),
    ssl.ca   = 'global-bundle.pem'
  )
}, error = function(e) {
  log_error(glue::glue("Database connection failed: {e$message}"), namespace = "app_data")
  NULL
})

if (is.null(CON)) {
  log_error("Database connection is NULL. Exiting.", namespace = "app_data")
  stop("Database connection failed. Exiting.")
} else {
  log_success("Database connection established successfully.", namespace = "app_data")
}

start_date <- params$start_date
end_date   <- params$end_date
start_dt   <- as.Date(start_date)
end_dt     <- as.Date(end_date)
```

```{r load-data}
source("mantua_payroll_data.R")
dbDisconnect(CON)
```

[*Report window: `r format(start_dt, "%B %d, %Y")` – `r format(end_dt, "%B %d, %Y")`*]{.display-7}

---

## 1 · Pay Summary by Firefighter

::: {.column-page}
```{r tbl-ff-summary}
ff_summary <- detail |>
  group_by(firefighter_id, full_name, firefighter_role, hourly_rate) |>
  summarise(
    incidents    = n_distinct(incident_id),
    responses    = n_distinct(response_id),
    actual_hours = sum(pmax(actual_hours, 0), na.rm = TRUE),
    paid_hours   = sum(paid_hours, na.rm = TRUE),
    paid_gross   = sum(paid_gross, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(paid_gross))

ff_summary |>
  select(-firefighter_id, -firefighter_role) |>
  gt() |>
  tab_header(
    title    = "Firefighter Pay — Summary",
    subtitle = glue::glue("{params$start_date}  →  {params$end_date}")
  ) |>
  cols_label(
    full_name    = "Name",
    hourly_rate  = "Rate ($/hr)",
    incidents    = "Incidents",
    responses    = "Responses",
    actual_hours = "Actual Hrs",
    paid_hours   = "Paid Hrs",
    paid_gross   = "Gross Pay"
  ) |>
  fmt_number(columns = c(actual_hours, paid_hours), decimals = 2) |>
  fmt_currency(columns = c(hourly_rate, paid_gross)) |>
  grand_summary_rows(
    columns = c(actual_hours, paid_hours, paid_gross),
    fns     = list(TOTAL = ~sum(., na.rm = TRUE)),
    fmt     = list(
      ~ fmt_number(., columns = c(actual_hours, paid_hours), decimals = 2),
      ~ fmt_currency(., columns = paid_gross)
    )
  ) |>
  tab_style(
    style     = cell_fill(color = "#f0f4f8"),
    locations = cells_grand_summary()
  ) |>
  opt_interactive(use_search = TRUE, use_sorting = TRUE, use_pagination = TRUE, page_size_default = 20)
```
:::
---

## 2 · Pay by Dispatch Type

::: {.column-page}
```{r tbl-dispatch}
detail |>
  group_by(dispatch_type, dispatch_code) |>
  summarise(
    incidents    = n_distinct(incident_id),
    responses    = n_distinct(response_id),
    actual_hours = sum(pmax(actual_hours, 0), na.rm = TRUE),
    paid_hours   = sum(paid_hours, na.rm = TRUE),
    paid_gross   = sum(paid_gross, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(paid_gross)) |>
  gt() |>
  tab_header(title = "Pay by Dispatch Type") |>
  cols_label(
    dispatch_type = "Dispatch Type",
    dispatch_code = "Dispatch Code",
    incidents     = "Incidents",
    responses     = "Responses",
    actual_hours  = "Actual Hrs",
    paid_hours    = "Paid Hrs",
    paid_gross    = "Gross Pay"
  ) |>
  fmt_number(columns = c(actual_hours, paid_hours), decimals = 2) |>
  fmt_currency(columns = paid_gross) |>
  opt_interactive(use_search = TRUE, use_sorting = TRUE, use_pagination = TRUE, page_size_default = 20)
```
:::
---

## 3 · Audit: Minimum Hour Adjustments

Responses where the minimum hour rule added time above actual hours worked.

::: {.column-page}
```{r tbl-min-hour}
min_adj <- detail |>
  filter(paid_hours > actual_hours) |>
  select(
    cad_identifier, dispatch_code,
    response_start, response_end,
    full_name, actual_hours, paid_hours,
    hourly_rate, actual_gross, paid_gross
  ) |>
  mutate(
    response_start = format(response_start, "%Y-%m-%d %H:%M"),
    response_end   = format(response_end,   "%Y-%m-%d %H:%M")
  )

if (nrow(min_adj) == 0) {
  cat("> ✅ No minimum hour adjustments in this window.")
} else {
  min_adj |>
    gt() |>
    tab_header(title = "Minimum Hour Adjustments") |>
    cols_label(
      cad_identifier = "CAD ID",
      dispatch_code  = "Dispatch Code",
      response_start = "Start",
      response_end   = "End",
      full_name      = "Firefighter",
      actual_hours   = "Actual Hrs",
      paid_hours     = "Paid Hrs",
      hourly_rate    = "Rate",
      actual_gross   = "Actual Pay",
      paid_gross     = "Paid Pay"
    ) |>
    fmt_number(columns = c(actual_hours, paid_hours), decimals = 2) |>
    fmt_currency(columns = c(hourly_rate, actual_gross, paid_gross)) |>
    opt_interactive(use_search = TRUE, use_sorting = TRUE, use_pagination = TRUE, page_size_default = 25)
}
```
:::
---

## 4 · Audit: Manual Time Adjustments

Responses where a manual time adjustment was applied.

::: {.column-page}
```{r tbl-time-adj}
time_adj <- detail |>
  filter(time_adjustment != 0) |>
  select(
    cad_identifier, dispatch_code,
    response_start, response_end,
    full_name, time_adjustment,
    actual_hours, paid_hours, hourly_rate, paid_gross
  ) |>
  mutate(
    response_start = format(response_start, "%Y-%m-%d %H:%M"),
    response_end   = format(response_end,   "%Y-%m-%d %H:%M")
  )

if (nrow(time_adj) == 0) {
  cat("> ✅ No manual time adjustments in this window.")
} else {
  time_adj |>
    gt() |>
    tab_header(title = "Line-Item Audit — Time Adjusted Responses") |>
    cols_label(
      cad_identifier  = "CAD ID",
      dispatch_code   = "Dispatch Code",
      response_start  = "Start",
      response_end    = "End",
      full_name       = "Firefighter",
      time_adjustment = "Adj (Hrs)",
      actual_hours    = "Actual Hrs",
      paid_hours      = "Paid Hrs",
      hourly_rate     = "Rate",
      paid_gross      = "Gross Pay"
    ) |>
    fmt_number(columns = c(time_adjustment, actual_hours, paid_hours), decimals = 2) |>
    fmt_currency(columns = c(hourly_rate, paid_gross)) |>
    opt_interactive(use_search = TRUE, use_sorting = TRUE, use_pagination = TRUE, page_size_default = 25)
}
```
:::
---

## 5 · Audit: Missing Pay Rates

Firefighters who responded in this window but have no entry in `firefighter_pay_rate`. They show as `$0.00` throughout — confirm this is intentional.

::: {.column-page}
```{r tbl-no-rate}
no_rate <- detail |>
  filter(hourly_rate == 0) |>
  distinct(firefighter_id, full_name, firefighter_role)

if (nrow(no_rate) == 0) {
  cat("> ✅ All responding firefighters have a pay rate on file.")
} else {
  no_rate |>
    gt() |>
    tab_header(title = "⚠️ Firefighters with No Pay Rate") |>
    tab_style(style = cell_fill(color = "#fff3cd"), locations = cells_body())
}
```
:::
---

## 6 · Audit: Zero or Negative Hour Responses

Responses where computed hours are `<= 0`, typically caused by a missing or incorrect `response_end`. These contribute `$0` to pay and should be corrected in the system.

::: {.column-page}
```{r tbl-bad-hours}
bad_hours <- detail |>
  filter(actual_hours <= 0 | is.na(actual_hours)) |>
  select(
    incident_id, cad_identifier, response_id,
    response_start, response_end,
    full_name, time_adjustment, actual_hours
  )

if (nrow(bad_hours) == 0) {
  cat("> ✅ No zero or negative hour records found.")
} else {
  bad_hours |>
    gt() |>
    tab_header(title = "⚠️ Zero / Negative Hour Responses") |>
    tab_style(style = cell_fill(color = "#f8d7da"), locations = cells_body())
}
```
:::
---

## 7 · Audit: Unusually Long Responses

Responses exceeding **12 hours**. May be legitimate major incidents or data entry errors — worth a quick look either way.

::: {.column-page}
```{r tbl-long}
long_resp <- detail |>
  mutate(base_hours = as.numeric(difftime(response_end, response_start, units = "hours"))) |>
  filter(base_hours > 12) |>
  select(
    incident_id, cad_identifier,
    response_start, response_end,
    full_name, base_hours, time_adjustment, paid_gross
  ) |>
  arrange(desc(base_hours))

if (nrow(long_resp) == 0) {
  cat("> ✅ No responses longer than 12 hours found.")
} else {
  long_resp |>
    gt() |>
    tab_header(title = "⚠️ Responses Exceeding 12 Hours") |>
    fmt_number(columns = c(base_hours, time_adjustment), decimals = 2) |>
    fmt_currency(columns = paid_gross) |>
    tab_style(style = cell_fill(color = "#fff3cd"), locations = cells_body())
}
```
:::
---

*Generated:* `r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`
